---
title: "Lecture 8: Sequential games"
date: "Week 3"
author: "Stuart J Morrison"
output:
      revealjs::revealjs_presentation:
        theme: solarized
        transition: none
        fig_width: 12
        fig_height: 5.5
---

<head>
  <link rel="stylesheet" href="styles.css">
</head>

# Letting someone go first

- Up to now, we've thought of games played at the same moment

- But now we'll think of different settings where one player goes first, and another player has to respond

$\quad\quad$ - First firm to enter a market might get higher market share

$\quad\quad$ - In tennis, the server gets to dictate how the first part of the rally is played

$\quad\quad$ - In chess, white gets to influence how the game will play out

- Whoever goes first might get an **advantage** because they can force the hand of the other play

# Letting someone go first

- In simultaneous games, each player has a best response to the other player

- In sequential games, only the second mover responds to the first

- The first mover can then predict what the second mover will do, and then choose its action based on that (backwards induction or backsolving)

# Let's go back to our simple entry game

There a two firms looking to enter a market. If both enter, they both make \$-1 profit, if only one enters then that firm makes \$1 profit, otherwise they both make no money

<table style="border-collapse: collapse;">
  <tr>
    <td style="border-right: 3px solid black; border-bottom: 3px solid black; padding: 30px;"></td>
    <td style="border-right: 1px solid black; border-bottom: 3px solid black; padding: 30px;">Firm 1 enters</td>
    <td style="border-bottom: 3px solid black; padding: 30px;">Firm 1 does not enter</td>
  </tr>
  <tr>
    <td style="border-right: 3px solid black; border-bottom: 1px solid black; padding: 30px;">Firm 2 enters</td>
    <td style="border-right: 1px solid black; border-bottom: 1px solid black; padding: 30px;">F1 profit = -1 <br> F2 profit = -1</td>
    <td style="border-bottom: 1px solid black; padding: 30px;">F1 profit = 0 <br> F2 profit = 1</td>
  </tr>
  <tr>
    <td style="border-right: 3px solid black; padding: 30px;">Firm 2 does not enter</td>
    <td style="border-right: 1px solid black; padding: 30px;">F1 profit = 1 <br> F2 profit = 0</td>
    <td style="padding: 30px;">F1 profit = 0 <br> F2 profit = 0</td>
  </tr>
</table>

- We showed in the last lecture, than when this is played simultaneously, each firm chooses with 50% probability to enter or not

- This is because the firms make simultaneous decisions, so neither knows what the other is going to do

# Now let's play the firm entry game as a sequential game

- But what if Firm 1 is quicker than Firm 2?

- If Firm 1 can choose before Firm 2 chooses - does it get an advantage in entering the market?

- We want to think about how Firm 1 choosing first changes the game outcome

# Now let's play the firm entry game as a sequential game

We draw this out in what's called a game tree - or in extensive form

```{r echo=FALSE,warning=FALSE,error=FALSE,message=FALSE,cache=FALSE}
  library(igraph)
library(ggraph)
library(ggplot2)
library(dplyr)

# Create the game tree structure
edges <- data.frame(
  from = c("F1", "F1", "F2_enter", "F2_enter", "F2_not_enter", "F2_not_enter"),
  to = c("F2_enter", "F2_not_enter", "(-1,-1)", "(1,0)", "(0,1)", "(0,0)"),
  action = c("Enter", "Not Enter", "Enter", "Not Enter", "Enter", "Not Enter")
)

# Create graph
g <- graph_from_data_frame(edges, directed = TRUE)

# Add node attributes
V(g)$type <- ifelse(V(g)$name %in% c("(-1,-1)", "(1,0)", "(0,1)", "(0,0)"), 
                    "terminal", "decision")
V(g)$player <- case_when(
  V(g)$name == "F1" ~ "Firm 1",
  V(g)$name %in% c("F2_enter", "F2_not_enter") ~ "Firm 2",
  TRUE ~ ""
)

# Plot the game tree
ggraph(g, layout = "tree") +
  geom_edge_link(aes(label = action), 
                 angle_calc = "along",
                 label_dodge = unit(2, "mm"),
                 arrow = arrow(length = unit(2, "mm")), size = 6) +
  geom_node_point(aes(color = type, size = type)) +
  geom_node_text(aes(label = ifelse(type == "terminal", name, player)),
  label_dodge = unit(-2, "mm"),
                 vjust = ifelse(V(g)$type == "terminal", -0.5, 0.5),
                 nudge_y = -0.2,
                 size = 6) +
  scale_color_manual(values = c("decision" = "lightblue", "terminal" = "lightcoral")) +
  scale_size_manual(values = c("decision" = 4, "terminal" = 3)) +
  theme_graph() +
  theme(legend.position = "none")
  
```

# Now let's play the firm entry game as a sequential game

We solve sequential games from the end to the start - called backwards induction

What would Firm 2 want to do at each node? 

Predicting what Firm 2 is going to do - what should Firm 1 do?


```{r echo=FALSE,warning=FALSE,error=FALSE,message=FALSE,cache=FALSE}
  library(igraph)
library(ggraph)
library(ggplot2)
library(dplyr)

# Create the game tree structure
edges <- data.frame(
  from = c("F1", "F1", "F2_enter", "F2_enter", "F2_not_enter", "F2_not_enter"),
  to = c("F2_enter", "F2_not_enter", "(-1,-1)", "(1,0)", "(0,1)", "(0,0)"),
  action = c("Enter", "Not Enter", "Enter", "Not Enter", "Enter", "Not Enter")
)

# Create graph
g <- graph_from_data_frame(edges, directed = TRUE)

# Add node attributes
V(g)$type <- ifelse(V(g)$name %in% c("(-1,-1)", "(1,0)", "(0,1)", "(0,0)"), 
                    "terminal", "decision")
V(g)$player <- case_when(
  V(g)$name == "F1" ~ "Firm 1",
  V(g)$name %in% c("F2_enter", "F2_not_enter") ~ "Firm 2",
  TRUE ~ ""
)

# Plot the game tree
ggraph(g, layout = "tree") +
  geom_edge_link(aes(label = action), 
                 angle_calc = "along",
                 label_dodge = unit(2, "mm"),
                 arrow = arrow(length = unit(2, "mm")), size = 6) +
  geom_node_point(aes(color = type, size = type)) +
  geom_node_text(aes(label = ifelse(type == "terminal", name, player)),
                nudge_y = -0.2,
                 vjust = ifelse(V(g)$type == "terminal", -0.5, 0.5),
                 size = 6) +
  scale_color_manual(values = c("decision" = "lightblue", "terminal" = "lightcoral")) +
  scale_size_manual(values = c("decision" = 4, "terminal" = 3)) +
  theme_graph() +
  theme(legend.position = "none")
  
```

# Now let's play the firm entry game as a sequential game

In a weird convention, we write our equilibria strategy out for every node from left to right for each layer.

So the equilibrium strategy for this game is:
\begin{equation*}\begin{aligned}
  \left( \underbrace{\text{Enter}}_{\text{Firm 1 node}}, \left(\underbrace{\text{Not enter}}_{\text{Firm 2 left node}}, \underbrace{\text{Enter}}_{\text{Firm 2 right node}}  \right) \right)
\end{aligned}\end{equation*}

We call this the **Subgame perfect equilibrium**

I think it's usually more useful to just think of the equilibrium outcome - ie, Firm 1 enters, Firm 2 does not enter

# Drawing out a more complex game

Last week I spoke about pharmaceutical companies breaking exclusivity restrictions. Let's map it out as a more complex sequential game.

- First, Firm 1 chooses to enter or not. They should be legally protected by an exclusivity provision.

- Then, Firm 2 chooses to enter or not. If they enter when Firm 1 has already entered, they are breaking the law.

- If both Firm 1 and 2 have entered - Firm 1 can choose to sue Firm 2.

- If Firm 1 acts as a monopoly, it earns \$5. If Firm 2 acts as a monopoly they earn \$3.

- If Firm 1 and 2 compete, they earn \$4 and \$2.

- If Firm 1 sues Firm 2: Firm 1 gets \$0.5 after legal fees, and Firm 2 loses an addition \$1.

