---
title: "Lecture 1: Competitive markets and welfare"
date: "Week 1"
author: "Stuart J Morrison"
output:
      revealjs::revealjs_presentation:
        theme: solarized
        transition: none
        fig_width: 12
        fig_height: 5.5
---

<head>
  <link rel="stylesheet" href="styles.css">
</head>

# Competitive markets

- From society's perspective, competitive markets are the best way to allocate goods and services between people

- As we learn about imperfect competition for the rest of this course - we'll use competitive markets as a benchmark 

- Requires unrealistic assumptions for a market to actually be 'perfectly competitive'...
        
- ...But some markets get close enough - like stock exchanges


# Remembering Competitive markets with the NYSE

- At each moment, people who want to sell stocks tell the NYSE their minimum price they'd take to sell their stock - their `willingness to sell'

- People who want to buy stocks tell the NYSE their maximum price they'd buy for - their `willingness to pay'

- The NYSE does two important things:

↬ matches buyers with the highest willingness to pay with sellers with the lowest willingness to sell - lets these people trade

↬ sets a price at which everyone trades

# Remembering Competitive markets with the NYSE

We can turn our share market into supply and demand

```{r echo=FALSE,warning=FALSE,error=FALSE,message=FALSE,cache=FALSE}
        
        library(tidyverse)

        library(ggplot2)

# Create data for increasing step function (supply)
supply_data <- tibble(
  shares = rep(0:10, each = 2),
  price = c(0,0,1, 1, 2, 2, 3, 3, 4, 4, 5, 5, 6, 6, 7, 7, 8, 8, 9, 9, 10, 10)
)

# Create data for decreasing step function (demand)  
demand_data <- tibble(
  shares = rep(0:10, each = 2),
  price = c(11,11,10, 10, 9, 9, 8, 8, 7, 7, 6, 6, 5, 5, 4, 4, 3, 3, 2, 2, 1, 1)
)

# Find intersection point (where supply price = demand price)
# This occurs around shares = 5.5, price = 5.5
intersection_shares <- 6
intersection_price <- 5.5

# Create the plot
ggplot() +
  # Supply curve (increasing, dark blue)
  geom_step(data = supply_data, aes(x = shares, y = price), 
            color = "darkblue", size = 1.2, direction = "hv") +
  
  # Demand curve (decreasing, dark red)
  geom_step(data = demand_data, aes(x = shares, y = price), 
            color = "darkred", size = 1.2, direction = "hv") +
  
  # Horizontal dotted line at intersection
  geom_hline(yintercept = intersection_price, 
             linetype = "dotted", color = "black", size = 0.8) +
  
  # Vertical dotted line at intersection
  geom_vline(xintercept = intersection_shares, 
             linetype = "dotted", color = "black", size = 0.8) +
  
  # Add text labels for the dotted lines
  annotate("text", x = 1, y = intersection_price + 0.3, 
           label = "Price at which everyone trades ($5.50)", 
           hjust = 0.5, vjust = 0, size = 3.5) +
  
  annotate("text", x = intersection_shares + 0.1, y = 0, 
           label = "Number of shares sold (6)", 
           hjust = 0, vjust = 0.5, size = 3.5, angle = 90) +
  
 annotate("text", x = 1.2, y = 0.5, label = "Willingness to sell (Supply curve)", hjust = 0, vjust = 0.5, size = 3.5, col = "darkblue") +
 
 annotate("text", x = 1.2, y = 10.5, label = "willingness to pay (Demand curve)", hjust = 0, vjust = 0.5, size = 3.5, col = "darkred") +

  # Customize axes and labels
  labs(x = "Number of shares", 
       y = "Price of the shares ($)") +
  
  # Set axis limits and breaks
  scale_x_continuous(limits = c(0, 11), breaks = 0:11) +
  scale_y_continuous(limits = c(0, 11), breaks = 0:10) +
  
  # Apply a clean theme
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )
        
```

# Remembering Competitive markets with the NYSE

What is the willingness to pay and willingness to sell for the 7th share? Will they trade?

```{r echo=FALSE,warning=FALSE,error=FALSE,message=FALSE,cache=FALSE}
        
        library(tidyverse)

        library(ggplot2)

# Create data for increasing step function (supply)
supply_data <- tibble(
  shares = rep(0:10, each = 2),
  price = c(0,0,1, 1, 2, 2, 3, 3, 4, 4, 5, 5, 6, 6, 7, 7, 8, 8, 9, 9, 10, 10)
)

# Create data for decreasing step function (demand)  
demand_data <- tibble(
  shares = rep(0:10, each = 2),
  price = c(11,11,10, 10, 9, 9, 8, 8, 7, 7, 6, 6, 5, 5, 4, 4, 3, 3, 2, 2, 1, 1)
)

# Find intersection point (where supply price = demand price)
# This occurs around shares = 5.5, price = 5.5
intersection_shares <- 6
intersection_price <- 5.5

# Create the plot
ggplot() +
  # Supply curve (increasing, dark blue)
  geom_step(data = supply_data, aes(x = shares, y = price), 
            color = "darkblue", size = 1.2, direction = "hv") +
  
  # Demand curve (decreasing, dark red)
  geom_step(data = demand_data, aes(x = shares, y = price), 
            color = "darkred", size = 1.2, direction = "hv") +
  
  # Horizontal dotted line at intersection
  geom_hline(yintercept = intersection_price, 
             linetype = "dotted", color = "black", size = 0.8) +
  
  # Vertical dotted line at intersection
  geom_vline(xintercept = intersection_shares, 
             linetype = "dotted", color = "black", size = 0.8) +
  
  # Add text labels for the dotted lines
  annotate("text", x = 1, y = intersection_price + 0.3, 
           label = "Price at which everyone trades ($5.50)", 
           hjust = 0.5, vjust = 0, size = 3.5) +
  
  annotate("text", x = intersection_shares + 0.1, y = 0, 
           label = "Number of shares sold (6)", 
           hjust = 0, vjust = 0.5, size = 3.5, angle = 90) +
  
 annotate("text", x = 1.2, y = 0.5, label = "Willingness to sell (Supply curve)", hjust = 0, vjust = 0.5, size = 3.5, col = "darkblue") +
 
 annotate("text", x = 1.2, y = 10.5, label = "willingness to pay (Demand curve)", hjust = 0, vjust = 0.5, size = 3.5, col = "darkred") +


  # Customize axes and labels
  labs(x = "Number of shares", 
       y = "Price of the shares ($)") +
  
  # Set axis limits and breaks
  scale_x_continuous(limits = c(0, 11), breaks = 0:11) +
  scale_y_continuous(limits = c(0, 11), breaks = 0:10) +
  
  # Apply a clean theme
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )
        
```

# Remembering Competitive markets with the NYSE

How much better off are the people that sell the first share?

```{r echo=FALSE,warning=FALSE,error=FALSE,message=FALSE,cache=FALSE}
        
        library(tidyverse)

        library(ggplot2)

# Create data for increasing step function (supply)
supply_data <- tibble(
  shares = rep(0:10, each = 2),
  price = c(0,0,1, 1, 2, 2, 3, 3, 4, 4, 5, 5, 6, 6, 7, 7, 8, 8, 9, 9, 10, 10)
)

# Create data for decreasing step function (demand)  
demand_data <- tibble(
  shares = rep(0:10, each = 2),
  price = c(11,11,10, 10, 9, 9, 8, 8, 7, 7, 6, 6, 5, 5, 4, 4, 3, 3, 2, 2, 1, 1)
)

# Find intersection point (where supply price = demand price)
# This occurs around shares = 5.5, price = 5.5
intersection_shares <- 6
intersection_price <- 5.5

# Create the plot
ggplot() +
  # Supply curve (increasing, dark blue)
  geom_step(data = supply_data, aes(x = shares, y = price), 
            color = "darkblue", size = 1.2, direction = "hv") +
  
  # Demand curve (decreasing, dark red)
  geom_step(data = demand_data, aes(x = shares, y = price), 
            color = "darkred", size = 1.2, direction = "hv") +
  
  # Horizontal dotted line at intersection
  geom_hline(yintercept = intersection_price, 
             linetype = "dotted", color = "black", size = 0.8) +
  
  # Vertical dotted line at intersection
  geom_vline(xintercept = intersection_shares, 
             linetype = "dotted", color = "black", size = 0.8) +
  
  # Add text labels for the dotted lines
  annotate("text", x = 1, y = intersection_price + 0.3, 
           label = "Price at which everyone trades ($5.50)", 
           hjust = 0.5, vjust = 0, size = 3.5) +
  
  annotate("text", x = intersection_shares + 0.1, y = 0, 
           label = "Number of shares sold (6)", 
           hjust = 0, vjust = 0.5, size = 3.5, angle = 90) +
  
 annotate("text", x = 1.2, y = 0.5, label = "Willingness to sell (Supply curve)", hjust = 0, vjust = 0.5, size = 3.5, col = "darkblue") +
 
 annotate("text", x = 1.2, y = 10.5, label = "willingness to pay (Demand curve)", hjust = 0, vjust = 0.5, size = 3.5, col = "darkred") +


  # Customize axes and labels
  labs(x = "Number of shares", 
       y = "Price of the shares ($)") +
  
  # Set axis limits and breaks
  scale_x_continuous(limits = c(0, 11), breaks = 0:11) +
  scale_y_continuous(limits = c(0, 11), breaks = 0:10) +
  
  # Apply a clean theme
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )
        
```

# Remembering Competitive markets with the NYSE

In total, how much better off is everyone that sells a share? How much better off is everyone that buys a share? In total, how much better off is society for trading?

```{r echo=FALSE,warning=FALSE,error=FALSE,message=FALSE,cache=FALSE}
        
        library(tidyverse)

        library(ggplot2)

# Create data for increasing step function (supply)
supply_data <- tibble(
  shares = rep(0:10, each = 2),
  price = c(0,0,1, 1, 2, 2, 3, 3, 4, 4, 5, 5, 6, 6, 7, 7, 8, 8, 9, 9, 10, 10)
)

# Create data for decreasing step function (demand)  
demand_data <- tibble(
  shares = rep(0:10, each = 2),
  price = c(11,11,10, 10, 9, 9, 8, 8, 7, 7, 6, 6, 5, 5, 4, 4, 3, 3, 2, 2, 1, 1)
)

# Find intersection point (where supply price = demand price)
# This occurs around shares = 5.5, price = 5.5
intersection_shares <- 6
intersection_price <- 5.5

# Create the plot
ggplot() +
  # Supply curve (increasing, dark blue)
  geom_step(data = supply_data, aes(x = shares, y = price), 
            color = "darkblue", size = 1.2, direction = "hv") +
  
  # Demand curve (decreasing, dark red)
  geom_step(data = demand_data, aes(x = shares, y = price), 
            color = "darkred", size = 1.2, direction = "hv") +
  
  # Horizontal dotted line at intersection
  geom_hline(yintercept = intersection_price, 
             linetype = "dotted", color = "black", size = 0.8) +
  
  # Vertical dotted line at intersection
  geom_vline(xintercept = intersection_shares, 
             linetype = "dotted", color = "black", size = 0.8) +
  
  # Add text labels for the dotted lines
  annotate("text", x = 1, y = intersection_price + 0.3, 
           label = "Price at which everyone trades ($5.50)", 
           hjust = 0.5, vjust = 0, size = 3.5) +
  
  annotate("text", x = intersection_shares + 0.1, y = 0, 
           label = "Number of shares sold (6)", 
           hjust = 0, vjust = 0.5, size = 3.5, angle = 90) +
  
 annotate("text", x = 1.2, y = 0.5, label = "Willingness to sell (Supply curve)", hjust = 0, vjust = 0.5, size = 3.5, col = "darkblue") +
 
 annotate("text", x = 1.2, y = 10.5, label = "willingness to pay (Demand curve)", hjust = 0, vjust = 0.5, size = 3.5, col = "darkred") +


  # Customize axes and labels
  labs(x = "Number of shares", 
       y = "Price of the shares ($)") +
  
  # Set axis limits and breaks
  scale_x_continuous(limits = c(0, 11), breaks = 0:11) +
  scale_y_continuous(limits = c(0, 11), breaks = 0:10) +
  
  # Apply a clean theme
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )
        
```

# Features of a competitive market to remember

- Everyone trades at the same price: At the intersection of willingness to pay and willingness to sell

- That is, no one has market power

↬ For most of our examples, the willingness to sell is the `marginal cost' of producing a good - so we say Price = Marginal Cost 

- If you sum up how much better off all the buyers and sellers are from trading - this is the best we can ever possibly do!

# Social welfare

In this class, our measure of how well off society is from trade is from summing up how much better off buyers and sellers are from trading

- We call how much better off buyers are `Consumer surplus'

- We call how much better off sellers are `Producer surplus'

How much better off society is from trading is Total surplus = Consumer surplus + Producer surplus

# Quick example (2 volunteers)

- A trade makes society better off

- We don't care how that benefit is distributed between people

↬ This makes the math easier - obviously not true in cases such as social welfare or tax policy

# Another quick example (2 volunteers)

- A trade makes society better off

- But a redistribution decreases total welfare, even if one party is better off

- We call these deadweight losses - its a drag on society


## Combining this all  together

- This course is all about `imperfect' competition

- We are going to think about how well off society in the real world compared to the `perfectly competitive' benchmark

- Remember Ivy Leagure enrollment from the introduction lecture

↬ These universities use market power and decrease the number of enrollments (imperfect competition) - this decreases surplus by artificially decreasing `trade' of university placements

↬ Some people who would be willing to pay the price of admission still can't get in - even if the universities have the resources to allow them in!

↬ We could then think about much better off society would be if they opened their doors to more students - but also how much worse off these universities would be in terms of value
